Software architectures to integrate workflow engines in science gateways

### 2016-04-22 paper.tex
Global read-through, several fixes, and tentative author list.

\title{Software architectures for workflow integration in science gateways}

@@

\author{}

\address{}

\begin{abstract}
%% Text of abstract
  We investigate software architectures to integrate workflow engines
  in science gateways. Based on our experience with the development
  and operations of the CBRAIN and VIP science gateways, and our past
  involvement in the SHIWA project, we abstract the software
  components and interactions in 6 architectures. We use these
  architectures to classify 18 existing systems and we compare these
  architectures using metrics measuring integration effort,
  robustness, extensibility, scalability and other specific
  features. Results show the pros and cons of the different
  architectures. Tight integration and sub-tasking require the least
  amount of integration effort and are also the most robust. Most
  architectures perform the same in terms of extensibility. The pool
  model is the most scalable. Meta-workflows are only available in
  nested workflows and workflow import.
\end{abstract}

@@

Workflow engines are critical components of the ecosystem of tools and
services offered by science gateways to leverage distributed
infrastructures for the efficient execution of applications through
the web. Several software architectures can be adopted to integrate
workflow engines in science gateways, with important consequences on
the required development effort and on the characteristics of the
resulting system. This paper describes and compares such architectures
based on system-independent representations of their main components
and interactions.

Our analysis is fed by experience gathered from the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years. It also builds on the lessons learned from several
science gateway and workflow-related projects, including the SHIWA
Simulation Platform~\cite{terstyanszky2014enabling}.

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework based on the functional
interactions between the main software components involved. In
Section~\ref{sec:evaluation}, we introduce metrics to measure
integration effort, robustness, extensibility, scalability and other
specific features related to the architectures. We use these metrics
to evaluate and compare the 6 architectures. The paper closes on a
discussion and conclusion that highlight architectural recommendations
for science gateway and workflow engine developers.

@@

In the last decade, the e-Science workflow community has developed
workflow systems to help developers access distributed infrastructures
such as clusters, grids, clouds and web services. These efforts
resulted in tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012}, etc. Such workflow engines usually
describe applications in a specific language that offers parallel data
and control flow constructs, visual edition features, links with
domain-specific tool repositories, provenance recording, among other
features (see review in~\cite{deelman2009workflows}).
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

At the same time, specific toolboxes were emerging in various
scientific domains to facilitate the assembly of software components
in consistent pipelines. In neuroimaging, our primary domain of
interest, tools such as Nipype~\cite{gorgolewski2011nipype},
PSOM~\cite{bellec2012pipeline}, SPM~\cite{ashburner2011spm} and
FSL~\cite{Jenkinson2012782} provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to create tasks, handle
their dependencies, and execute them. For instance, FSL can launch
tasks on SGE through its fsl\_sub tool, Nipype has execution
plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
and PSOM is interfaced to Torque, SGE, MOAB, LSF and Condor. Some of
these tools also support advanced features such as provenance tracking
(Nipype, PSOM), redundancy detection across analyses to avoid
re-computation (PSOM), and so on. A wide-array of workflows have been
implemented using these pipelines and are now widely shared across
neuroimaging groups world-wide. \todo{Provide more context on that?}
These domain-specific engines represent a tremendous opportunity for
science gateways to leverage existing tools and applications. They
nicely complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

For the sake of the analysis conducted in this paper, we
define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
software that submits interdependent computing \emph{tasks} to an
infrastructure (local server, cluster, grid or cloud) based on a
workflow description (a.k.a \emph{workflow}) and using input data that
may be files, data base entries, or simple parameter values. Although
simplistic, this definition covers both e-Science and domain-specific
engines. Some workflow engines, usually e-Science ones, may transfer
data across the infrastructure, and others, usually domain-specific
ones, may leave it to external processes. Workflows may be expressed
in any language, including high-level XML or JSON dialects such as
Scufl or Hyperflow, and low-level scripts such as
Bash.

@@

Science gateways 
are used to share resources in a community and to provide performance
through facilitated access to storage and computing power. They are
usually accessible through a web interface that helps users manage
access rights, data transfers, task execution and authentication on
multiple computing and storage locations. Workflow engines are of
course part of this ecosystem as core components to implement and
execute applications.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways may be tailored to a particular
workflow engine that manages critical functions such as task
execution, data transfers and authentication. Other gateways may be
more general and host applications executed by different types of
engines. Extensibility is an important property of such integration:
new workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure may be covered.

Scalability is also an important concern, since science gateways are
multi-user systems, manage substantial amounts of data and execute
computationally-intensive analyses. For this purpose, science gateways
may include different instances of the same engine type and implement
some load-balancing policy among them. They may also enforce
scheduling policies on the infrastructure, e.g. to ensure fairness
among users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

Robustness is a key part of science gateway user experience. In this
respect, the multiplicity of software components and functions is a
challenge as it often leads to complex architectures that are
therefore prone to failure. To offer good user experience, science
gateways should remain as transparent as possible, which requires
robustness.

@@

Various science gateways have been developed in the last years,
including science gateway frameworks used as building blocks to
assemble science gateways. The main frameworks are
WS-PGRADE/gUSE~\cite{Kacsuk2012}, the Catania Science Gateway
Framework~\cite{Ardizzone2012} and Apache
Airavata~\cite{marru2011apache}. Numerous science gateway instances
were built using such frameworks, in various disciplines. Other
gateway instances were also built independently from any framework,
using regular web development tools and various libraries to manage
distributed systems. CBRAIN and VIP are two of such science gateways
that were initially developed for medical imaging analysis and from
which we gathered experience about the architectures described in
this paper.

@@

The Virtual Imaging Platform
(VIP\footnote{\url{http://www.creatis.insa-lyon.fr/vip}})~\cite{GLAT-13}
is a web portal for medical simulation and image data analysis. It
leverages computing and storage resources available in the biomed
Virtual Organization of the European Grid Infrastructure (EGI) to
offer an open service to academic researchers worldwide. VIP
completely masks the infrastructure to enable a transparent user
experience. It uses the MOTEUR workflow engine~\cite{GLAT-13} and the
DIRAC pilot-job framework to execute applications on EGI. Almost 900
users were registered in VIP in April 2016 and the average CPU
consumption between January 2013 and March 2016 was 35 CPU years per
month.

The architectures described hereafter also build on experience with
the SHIWA Simulation Platform~\cite{terstyanszky2014enabling}. The
SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
developed in European projects
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}. It focuses on workflow
interoperability for various disciplines, in particular through the
building of meta-workflows, i.e. workflows that are built out of other
workflows. It implements the so-called Coarse-Grained Interoperability
framework where workflow engines are integrated in the science gateway
as other applications and invoked through a dedicated service. The
Fine-Grained Interoperability framework was also developed in SHIWA,
where conversions between workflow languages are achieved through a
common intermediate
representation~\cite{plankensteiner-prodan-etal:2013}. \note{Tristan}{Silvia,
  do you think it is legitimate to refer to SHIWA like this in the
  paper, or should we involve Gabor in the writing?}

@@

In this paper, infrastructure consists of the computing and storage
resources involved in the workflow execution, and the software
services used to access these resources. Infrastructures can be
servers, clusters, grids or clouds. Some workflow engines and science
gateways may assume specific characteristics about the infrastructure,
such as the presence of a shared file system between the computing
nodes, the availability of a global task meta-scheduler, the presence
of a file catalog, etc. In the analysis presented below, such specific
functions are not discussed. Instead, we consider the infrastructure
as an abstract system that can execute tasks and store files
regardless of the enabling mechanisms.

@@

The architectures are diagrammed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software \emph{components} and \emph{interactions}. Software
components include science gateway and infrastructure in all
architectures while workflow service, workflow pool and agent are
involved in some architectures. The workflow engine itself is
represented by a specific symbol with dotted lines. Software
interactions among these components are represented with arrows.
\emph{Abstract} interactions are specific types of interactions that
may be implemented by various different software interactions. They
are represented with dotted arrows and labeled with \texttt{*}. An
architecture that has an abstract interaction is an abstract
architecture.  Red color is used to represent the components and
interactions that are specific to workflow execution, i.e. that are
not involved in any other process. Tasks refer to computing tasks that
are created by the workflow engine and executed on the
infrastructure. Data represents any type of file, or database that is
involved in the workflow execution and stored on the infrastructure
while the workflow executes. It does not cover workflow parameters
such as strings, numbers, etc.

The remainder of this section describes the different types of
interactions involved in the architectures, details architectures and
illustrates them using examples of real
systems. Table~\ref{table:system-classification} summarizes the
classification of a few systems by architecture.

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Sub-tasking]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}

@@

\begin{enumerate}[leftmargin=0cm,itemindent=0.6cm,label=\texttt{(\alph*)}]

@@

  the infrastructure, which may or may not require administrative
  privileges on the infrastructure; (\texttt{a$_2$}) the workflow is
  configured in the science gateway so that it becomes available to
  users. Note that integrating a workflow is not the same process as
  integrating a workflow \emph{engine}.

@@

  or a cloud, it is for instance implemented using libraries
  implementing standards such as SAGA~\cite{goodale2006saga},
  DRMAA~\cite{troger2012distributed} or OCCI~\cite{edmonds2012toward}.

@@

\item Workflow control: operations to execute a workflow with an

@@

  etc. Workflow control can be coarse-grained (a.k.a. black box) or

@@

\item Pool-agent: specific to the architecture described
  in Section~\ref{sec:pool}. This is an interaction used when agents retrieve
  work from a central pool. It covers agent registration and

@@

  the agent, mechanisms to update work status, and so on. 

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored. In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in LONI Pipeline
Environment~\cite{dinov2009efficient}, in the Distributed application
runtime environment (DARE~\cite{maddineni2012distributed}), and in the
Catania Science Gateway Framework~\cite{Ardizzone2012} which is used
for instance in the DECIDE gateway for 
neuroimaging~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}). 
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

\begin{figure*}
\centering
\def\svgwidth{1.5\columnwidth}
\input{figures/VIP.pdf_tex}
\caption{Architecture of the Virtual Imaging Platform that integrates
  the MOTEUR engine through service invocation.}
\label{fig:vip-architecture}
\end{figure*}

@@

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the system
in~\cite{wu2010accelerating}, the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012} and the numerous science gateway instances
that rely on WS-PGRADE/gUSE, for instance
e-BioInfra~\cite{shahand2015data}, MoSGrid~\cite{krüger2014mosgrid},
and many others~\cite{kacsuk2014science}. The
integration between the HubZero science gateway and Pegasus workflow
engine performed in~\cite{CPE:CPE3257} also use a service architecture
where the \texttt{d} interaction is implemented as a set of calls to
the Pegasus command-line tools (e.g. pegasus-status,
pegasus-statistics, etc). Figure~\ref{fig:vip-architecture} shows the
architecture of the Virtual Imaging Platform, where the MOTEUR
workflow engine~\cite{GLAT-08i} is invoked as a service at step
\texttt{3} (interaction \texttt{d} on
Figure~\ref{fig:architectures}). Step \texttt{2} on
Figure~\ref{fig:vip-architecture} corresponds to interaction
\texttt{c$_1$}, and step \texttt{4} maps to interactions \texttt{b}
and \texttt{c$_2$} (in VIP, workflow data transfers are performed by
the tasks and embedded in their descriptions).

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is a particular
task that can submit sub-tasks to the science gateway through
interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks but their execution is delegated to
the science gateway that executes them on the infrastructure through
interaction \texttt{b}. Although the science gateway has no global
vision of the workflow, it can keep track of the sub-tasks submitted
by a given task, for instance to be able to cancel them when the task
is canceled. The science gateway may also implement mechanisms to
facilitate the handling of task dependencies, for instance basic
dependency lists as available in most batch managers (see for instance
attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}}\\

>>>>

\title{Software architectures to integrate workflow engines in science gateways}

@@

\author[mcgill,creatis]{Tristan Glatard}
\author[mcgill]{Marc-Etienne Rousseau}
\author[creatis]{Sorina Camarasu-Pop}

\author[mcgill]{Reza Adalat}
\author[mcgill]{Natacha Beck}
\author[mcgill]{Najmeh Khalili-Mahani}
\author[criugm]{Pierre-Olivier Quirion}
\author[mcgill]{Pierre Rioux}

\author[criugm]{Pierre Bellec}
\author[mcgill]{Alan C. Evans}

\address[mcgill]{McConnell Brain Imaging Centre, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montreal, QC, Canada.}

\begin{abstract}
  We investigate software architectures to integrate workflow engines
  in science gateways. Based on our experience with the development
  and operations of the CBRAIN and VIP science gateways, and our past
  involvement in the SHIWA project, we abstract software components
  and interactions in 6 architectures that we use to classify 18
  systems. We compare these architectures using metrics measuring
  integration effort, robustness, extensibility, scalability and other
  specific features. Results show the pros and cons of the different
  architectures: tight integration and sub-tasking require the least
  amount of integration effort and are also the most robust, most
  architectures perform the same in terms of extensibility, the pool
  model is the most scalable and meta-workflows are only available in
  nested workflows and workflow import. These results provide insights
  to software architects on how to integrate science gateways and
  workflow engines.
\end{abstract}

@@

Workflow engines are critical components of the ecosystem of tools and
services offered by science gateways to leverage distributed
infrastructures for the efficient execution of applications through
the web. Several software architectures can be adopted to integrate
workflow engines in science gateways, with important consequences on
the required development effort and on the resulting system. This
paper describes, exemplifies and compares such architectures, based on
system-independent representations of their main components and
interactions.

Our analysis is fed by experience gathered from the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years. It also builds on the lessons learned from several
science gateway and workflow-related projects, including
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and related
initiatives.

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework based on the functional
interactions between the main software components involved. In
Section~\ref{sec:evaluation}, we introduce metrics to measure
integration effort, robustness, extensibility, scalability and other
specific features of the architectures. We use these metrics to
evaluate and compare the 6 architectures. The paper closes on a
discussion and conclusion that highlight architectural recommendations
for science gateway and workflow engine architects.

@@

In the last decade, the e-Science community has developed workflow
systems to help developers access distributed infrastructures such as
clusters, grids, clouds and web services. These efforts resulted in
tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and others. Such workflow engines usually
describe applications in a high-level language that offers specific
data and control flow constructs, parallelization operators, visual
edition features, links with domain-specific tool repositories,
provenance recording, among other features. An overview of workflow
systems features and capability is available
in~\cite{deelman2009workflows}.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to handle high-throughput
task executions. For instance, FSL can launch tasks on SGE through its
fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF and Condor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now widely shared across neuroimaging groups
world-wide. These
domain-specific engines represent a tremendous opportunity for science
gateways to leverage existing tools and applications. They nicely
complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

For the sake of the analysis conducted in this paper, we define a
\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science and
domain-specific engines. Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave it to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources in a community and to
provide performance through facilitated access to storage and
computing power. They are usually accessible through a web interface
that helps users manage access rights, data transfers, task execution
and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways may be tailored to a particular
engine that manages critical functions such as task execution, data
transfers and authentication. Other gateways may be more general and
host applications executed by different types of
engines. Extensibility is an important property of the integration, as
new workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted.

Scalability is also an important concern, since science gateways are
multi-user systems that manage substantial amounts of data and execute
computationally-intensive analyses. To this end, science gateways
may include different instances of the same engine type and implement
some load-balancing policy among them. They may also use advanced
task scheduling policies on the infrastructure, e.g. to ensure fairness
among users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

Robustness is a key part of science gateway user experience. In this
respect, the multiplicity of software components and interactions
involved in architectures is a challenge as it often leads to complex
systems that are therefore prone to failure. To offer good user
experience, science gateways should remain as transparent as possible,
which requires robustness.

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble science gateways. The main
frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
were built using such frameworks\note{Tristan}{Add references}. Other
gateway instances were also built independently from any framework,
using regular web development tools and libraries to access
distributed systems. CBRAIN and VIP are two of such science gateways,
from which we gathered experience about the architectures described in
this paper.

@@

The Virtual Imaging Platform
(VIP\footnote{\url{http://www.creatis.insa-lyon.fr/vip}})~\cite{GLAT-13}
is a web portal for medical simulation and image data analysis. It
leverages computing and storage resources available in the biomed
Virtual Organization of the European Grid Infrastructure (EGI) to
offer an open service to academic researchers worldwide. VIP
completely masks the infrastructure to enable a transparent user
experience. It uses the MOTEUR workflow engine~\cite{GLAT-13} and the
DIRAC pilot-job framework~\cite{tsaregorodtsev2008dirac} to execute
applications on EGI. Almost 900 users were registered in VIP as of
April 2016 and the average CPU consumption between January 2013 and
March 2016 was 35 CPU years per month.

The architectures described hereafter also build on experience with
the SHIWA Simulation Platform~\cite{terstyanszky2014enabling}. The
SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
developed in European projects
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}. It focuses on workflow
interoperability for various disciplines, in particular through the
building of meta-workflows, i.e. workflows that are built out of other
workflows. It implements the so-called Coarse-Grained Interoperability
framework where workflow engines are integrated in the science gateway
as other applications and invoked through a dedicated service. The
Fine-Grained Interoperability framework was also developed in the
SHIWA project, where conversions between workflow languages are
achieved through a common intermediate
representation~\cite{plankensteiner-prodan-etal:2013}. \note{Tristan}{Silvia,
  do you think it is legitimate to refer to SHIWA like this in the
  paper, or should we involve Gabor in the writing?}

@@

In this paper, infrastructure consists of the computing and storage
resources involved in the workflow execution, and the software
services used to access these resources. Infrastructures can be
computing or file servers, databases, clusters, grids or clouds. Some
workflow engines and science gateways may assume specific
characteristics about the infrastructure, such as the presence of a
shared file system between the computing nodes, the availability of a
global task meta-scheduler, the presence of a file catalog, etc. In
the analysis presented below, such specific functions are not
discussed. Instead, we consider the infrastructure as an abstract
system that can execute tasks and store data regardless of the
enabling mechanisms.

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in Figure~\ref{fig:notations}.

Architectures are described from their main software \emph{components}
and \emph{interactions}. Software components include science gateway
and infrastructure in all architectures while workflow service,
workflow pool and agent are involved in some architectures only. The
workflow engine itself is represented by a specific symbol with dotted
lines.

Software interactions among these components are represented with
arrows.  \emph{Abstract} interactions are specific types of
interactions that may be implemented using different components and
interactions. They are represented with dotted arrows and labeled with
\texttt{*}. An architecture that has an abstract interaction is an
abstract architecture that can be instantiated in various ways.

Red color is used to represent the components and interactions that
are specific to workflow execution, i.e. that are not involved in any
other process. Tasks refer to computing tasks created by the workflow
engine and executed on the infrastructure. Data represents any type of
file or database involved in workflow execution and stored on the
infrastructure. It does not cover workflow parameters such as strings,
numbers, etc.

The remainder of this section describes the different types of
interactions involved in the architectures. Architectures are then
detailed and illustrated using examples of real
systems. Table~\ref{table:system-classification} summarizes the
classification of a few systems by architecture.

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Sub-tasking]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012}, 
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012}, 
            Science gateways in ~\cite{kacsuk2014science}$^w$
            } \\
  Sub-task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$\todo{others}\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$
\end{tabular}
\caption{Classification of science gateway based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

\begin{enumerate}[leftmargin=0cm,itemindent=0.65cm,label=\texttt{(\alph*)}]

@@

  the infrastructure, which may require administrative privileges on
  the infrastructure; (\texttt{a$_2$}) the workflow is configured in
  the science gateway so that it becomes available to users. Note that
  integrating a workflow is not the same process as integrating a
  workflow \emph{engine}.

@@

  or a cloud, it may for instance be achieved using libraries that
  implement standards such as SAGA (Simple API for Grid
  Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
  Management Application API~\cite{troger2012distributed}) or OCCI
  (Open Cloud Computing Interface~\cite{edmonds2012toward}).

@@

\item Workflow control: operations to control workflow execution in an

@@

  etc. Workflow control can be coarse-grained (black box) or

@@

\item Pool-agent: specific to the pool architecture described in
  Section~\ref{sec:pool}. This is an interaction used when agents
  retrieve work from a central pool. It covers agent registration and

@@

  the agent, mechanisms to update work status, and so on. A similar
  type of interaction is used in pilot-job
  systems~\cite{turilli2015comprehensive} and other agent-based
  computing models.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored. In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment (DARE~\cite{maddineni2012distributed})
and in LONI Pipeline Environment~\cite{dinov2009efficient}.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

\begin{figure*}
\centering
\def\svgwidth{1.5\columnwidth}
\input{figures/VIP.pdf_tex}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  on Figure~\ref{fig:architectures}, step \texttt{2} corresponds to
  interaction \texttt{c$_1$}, and step \texttt{4} maps to interactions
  \texttt{b} and \texttt{c$_2$} (in VIP, workflow data transfers are
  performed by the tasks and embedded in their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the system
in~\cite{wu2010accelerating} and the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012}. The integration between the
HubZero science gateway and Pegasus workflow engine performed
in~\cite{CPE:CPE3257} also uses a service architecture where the
\texttt{d} interaction is implemented as a set of calls to Pegasus
command-line tools (e.g. pegasus-status, pegasus-statistics,
etc). Figure~\ref{fig:vip-architecture} shows the architecture of the
Virtual Imaging Platform, where the MOTEUR workflow
engine~\cite{GLAT-08i} is invoked through a service.

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is a particular
task that can submit sub-tasks to the science gateway through
interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated to
the science gateway that executes them on the infrastructure through
interaction \texttt{b}. Although the science gateway has no global
vision of the workflow, it can keep track of the sub-tasks submitted
by a given task, for instance to be able to cancel them when the task
is canceled. The science gateway may also implement mechanisms to
facilitate the handling of task dependencies, for instance basic
dependency lists as available in most batch managers (see for instance
attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
